Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-05-01 LaTeX/main_revised.tex
updated conclusions, but havent referenced them proprely

Neo-W alleles, on the other hand, can invade an ancestral XY system because, under some conditions, the neo-W can become more strongly specialized for females (the full invasion conditions are given in \nameref{app:inv_cond}; equations \ref{Binvasion} and \ref{Ainvasion}). 
Thus,
\vspace{0.5cm}

\noindent\textit{Conclusion 1A:}
%\textbf{selection on loci within the non-recombining region around the sex-determining locus ($r\approx0$) allows the invasion of less closely linked sex-determining alleles ($r<R\leq1/2$) during a trans-GSD transition} (XY $\leftrightarrow$ ZW).
\textbf{Selection on loci in or near the non-recombining region around the sex-determining locus ($r\approx0$) prevents cis-GSD transitions (XY $\leftrightarrow$ XY, ZW $\leftrightarrow$ ZW) but can spur trans-GSD transitions (XY $\leftrightarrow$ ZW), which weaken sex-linkage.}
\vspace{0.5cm}

Conclusion 1A does not depend on the form of selection maintaining a polymorphism (sexually-antagonistic selection, overdominance, ploidally-antagonistic selection, or some combination, Fig \ref{fig:SexAntagTighter}, \nameref{fig:temporalOverdominance}, \nameref{fig:regionPloidAntag}, and \nameref{fig:SexAntagTighterMaleDrive}).  
The conditions become more restrictive, however, with increasing recombination ($R$) between the new sex-determining locus and the selected locus.  
The invasion of completely unlinked neo-W alleles ($R = 1/2$) can occur with overdominance in males or with haploid selection (e.g., when male drive is opposed by diploid selection, as in \cite{Kozielska:2010vm}) but is not possible with only sexually-antagonistic selection where selection is directional in each diploid sex (see \nameref{file:Mathematica}).
Here, we focus on cases where there is no haploid selection (Fig \ref{fig:SexAntagTighter}A) and discuss the additional effect of haploid selection in \nameref{app:inv_cond} (see also Conclusion 1C). 
%Essentially, neo-W alleles are able to invade (even with $R>r$) because the ancestral X experiences selection in males. 
%This can lead to (1) XX females having lower than maximum fitness ($\Lambda_{W'A}^{(XY)}>1$) and/or (2) the Y carrying alleles that are beneficial in females.

@@

\noindent\textit{Conclusion 1B:}
\textbf{With tight linkage between a selected locus and the ancestral sex-determining locus ($r\approx0$), selection to better specialize on the ancestrally-non-heterogametic sex can drive trans-GSD transitions (XY $\leftrightarrow$ ZW) that weaken sex-linkage ($r<R\leq1/2$; potentially even shifting sex determination to a different chromosome) and sometimes lead to polymorphic sex-determination systems.}
\vspace{0.5cm}

Adding haploid selection (\nameref{app:inv_cond}) we also conclude that
\vspace{0.5cm}

\noindent\textit{Conclusion 1C:}
\textbf{
With tight linkage between a selected locus and the ancestral sex-determining locus ($r\approx0$), haploid selection allows trans-GSD transitions (XY $\leftrightarrow$ ZW) that weaken sex-linkage ($r<R\leq1/2$; potentially even shifting sex determination to a different chromosome) even when selection is sexually-antagonistic with directional selection in each diploid sex.  
These trans-GSD transitions equalize the sex ratio.
}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\noindent\textit{Conclusion 2A:}\
\textbf{New sex-determining alleles (causing cis- \textit{or} trans-GSD transitions) can spread if they arise morely closely linked with a locus that experiences selection than the ancestral-sex-determining locus is ($R<r$).} 
\vspace{0.5cm}

%However, with haploid selection and some ancestral sex-linkage ($r<1/2$; allowing allele frequency differences on the X and Y), the additional term in Eq \eqref{eq:lambda_neoW} can be positive.
%This implies, 
%\vspace{0.5cm}
\noindent\textit{Conclusion 2B:}\
\textbf{New sex-determining alleles (causing trans-GSD transitions, XY $\leftrightarrow$ ZW) can spread even if they arise less closely linked with a locus that experiences selection than the ancestral-sex-determining locus is ($r<R$), requiring haploid selection and some initial sex-linkage ($r<1/2$) as necessary conditions.} 
\vspace{0.5cm}

@@

\noindent\textit{Conclusion 3:}
\textbf{When selection is weak relative to recombination, the presence of haploid selection equally favors the spread of new sex determination systems that reduce sex-ratio bias (benefiting from Fisherian sex ratio selection) or generate a sex-ratio bias (benefiting from associations with selected alleles). } 
\vspace{0.5cm}
%\noindent\textcolor{red}{\textit{Conclusion 3:}
%\textbf{when selection is weak relative to recombination, sex-ratio selection is exactly equal in magnitude to the selection imposed by associations with selected alleles. } }
%%\textcolor{red}{[new result]}
%\vspace{0.5cm}

@@

\noindent\textit{Conclusion 4:}
\textbf{
Transitions from genetic to environmental sex-determination are not straightforwardly predicted by selection to balance the zygotic sex ratio when haploid selection is present. 
}
%\textcolor{red}{[new result]}

>>>>

Neo-W alleles, on the other hand, can invade an ancestral XY system (the full invasion conditions are given in \nameref{app:inv_cond}; equations \ref{Binvasion} and \ref{Ainvasion}).
Thus,
\vspace{0.5cm}

\noindent\textit{Conclusion 1:}
%\textbf{selection on loci within the non-recombining region around the sex-determining locus ($r\approx0$) allows the invasion of less closely linked sex-determining alleles ($r<R\leq1/2$) during a trans-GSD transition} (XY $\leftrightarrow$ ZW).
\textbf{Selection on loci in or near the non-recombining region around the sex-determining locus ($r\approx0$) prevents cis-GSD transitions (XY $\leftrightarrow$ XY, ZW $\leftrightarrow$ ZW) but can spur trans-GSD transitions (XY $\leftrightarrow$ ZW).}
\textcolor{red}{[Not clear from van Doorn \& Kirkpatrick 2007, 2010]}
\vspace{0.5cm}

\noindent Trans-GSD transitions can occur because, under some conditions, the neo-W can become more strongly specialized for females.
This does not depend on the form of selection maintaining a polymorphism (sexually-antagonistic selection, overdominance, ploidally-antagonistic selection, or some combination, Fig \ref{fig:SexAntagTighter}, \nameref{fig:temporalOverdominance}, \nameref{fig:regionPloidAntag}, and \nameref{fig:SexAntagTighterMaleDrive}).

The conditions allowing trans-GSD transitions become more restrictive with increasing recombination ($R$) between the new sex-determining locus and the selected locus.  
The invasion of completely unlinked neo-W alleles ($R = 1/2$) can occur with overdominance in males or with haploid selection (e.g., when male drive is opposed by diploid selection, as in \cite{Kozielska:2010vm}) but is not possible with only sexually-antagonistic selection where selection is directional in each diploid sex (see \nameref{file:Mathematica}).
To clarify the conditions under which trans-GSD transitions occur, we focus here on cases where there is no haploid selection (Fig \ref{fig:SexAntagTighter}A) and discuss the additional effect of haploid selection in \nameref{app:inv_cond} (see also Conclusion 2B). 
%Essentially, neo-W alleles are able to invade (even with $R>r$) because the ancestral X experiences selection in males. 
%This can lead to (1) XX females having lower than maximum fitness ($\Lambda_{W'A}^{(XY)}>1$) and/or (2) the Y carrying alleles that are beneficial in females.

@@

\noindent\textit{Conclusion 2A:}
\textbf{
With tight linkage between a selected locus and the ancestral sex-determining locus ($r\approx0$), trans-GSD transitions (XY $\leftrightarrow$ ZW) can be driven by selection to better specialize on the ancestrally-non-heterogametic sex.
These transitions weaken sex-linkage ($r<R$), potentially shifting sex determination to a different chromosome ($R=1/2$). 
Such transitions can lead to polymorphic sex-determination systems.
}
\textcolor{red}{[consistent with special case of Kozielska et al 2010]}
\vspace{0.5cm}

\nameref{app:inv_cond}) shows that Conclusion 2A holds with or without haploid selection, and further demonstrates that
\vspace{0.5cm}

\noindent\textit{Conclusion 2B:}
\textbf{
With tight linkage between a selected locus and the ancestral sex-determining locus ($r\approx0$), haploid selection can drive trans-GSD transitions (XY $\leftrightarrow$ ZW) that weaken sex-linkage ($r<R$), even when selection is sexually-antagonistic with directional selection in each diploid sex.
These trans-GSD transitions are driven by selection to equalize the sex ratio.
As in Conclusion 2A, such transitions potentially shift sex determination to a different chromosome ($R=1/2$) and can lead to polymorphic sex-determination systems.}
\textcolor{red}{[consistent with special case of Kozielska et al 2010]}

@@

\vspace{0.5cm}
\noindent\textit{Conclusion 3A:}\
\textbf{New sex-determining alleles (causing cis- \textit{or} trans-GSD transitions) can spread if they arise more closely linked with a locus that experiences selection than the ancestral-sex-determining locus is ($R<r$).}
\textcolor{red}{[consistent with van Doorn \& Kirkpatrick 2007, 2010]} 
\vspace{0.5cm}

%However, with haploid selection and some ancestral sex-linkage ($r<1/2$; allowing allele frequency differences on the X and Y), the additional term in Eq \eqref{eq:lambda_neoW} can be positive.
%This implies, 
%\vspace{0.5cm}
\noindent\textit{Conclusion 3B:}\
\textbf{New sex-determining alleles (causing trans-GSD transitions, XY $\leftrightarrow$ ZW) can spread even if they arise less closely linked with a locus that experiences selection than the ancestral-sex-determining locus is ($r<R$), requiring haploid selection and some initial sex-linkage ($r<1/2$) as necessary conditions.}
\textcolor{red}{[consistent with examples in Ubeda et al 2015]}  
\vspace{0.5cm}

@@

\noindent\textit{Conclusion 4A:}
\textbf{When selection is weak relative to recombination, the presence of haploid selection equally favors the spread of new sex determination systems that reduce sex-ratio bias (benefiting from Fisherian sex ratio selection) or generate a sex-ratio bias (benefiting from associations with selected alleles). } 
\textcolor{red}{[new result]}  
\vspace{0.5cm}
%\noindent\textcolor{red}{\textit{Conclusion 3:}
%\textbf{when selection is weak relative to recombination, sex-ratio selection is exactly equal in magnitude to the selection imposed by associations with selected alleles. } }
%%\textcolor{red}{[new result]}
%\vspace{0.5cm}

@@

\noindent\textit{Conclusion 4B:}
\textbf{
Transitions from genetic to environmental sex-determination are not straightforwardly predicted by selection to balance the zygotic sex ratio when haploid selection is present. 
}
\textcolor{red}{[new result]}  
%\textcolor{red}{[new result]}
